A Systematic Literature Review on Extensions to Role Based Access Control

### 2013-04-02 paper/sections/abstract.tex
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions called roles.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: \textit{The goal of this work is to provide practioners an assessment of the state of extended models of RBAC and researchers with insights into the lack of robust evaluation of RBAC extension models.}
\\
\\
\noindent
{\bf Method}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}:
\\
\\
\noindent
{\bf Conclusions}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.

>>>>

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's by the National Institute of Standards and Technology (NIST),
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practioners and researchers in choosing an exentsion to the RBAC standard model, and in understanding
how extensions to RBAC are evaluated by providing an assessment of the state of extensions to the RBAC standard model through: establisment of a 
set of extension categories, an examination of the state of the art in evaluations of extension models and a breakdown of the reasoning and motivations
that have led to extension of the RBAC standard.
\\
\\
\noindent
{\bf Method}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. The review we performed yielded 1716 papers, of which 29 were deemed primary sources 
for inclusion as model extensions to the RBAC standard model. The papers we collected were from electronic digital libraries
IEEE, Google Scholar, CiteSeerX and ACM based on a set of exclusion and inclusion criteria.  
We performed a comparative analysis of the primary sources to find relationships among extended models and to analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}: Our results showed that extensions to RBAC can be classified under seven categories: Constraint, Context, 
Organization, Privacy, Resource, Task, Spatio-Temporal, Spatial and Temporal. We found a range of motivations behind RBAC extensions from 
lack of fine-grained controls to a need for context placed on permissions, roles and assignment.  We found that 12 models presented no evidence
of an evaluation, 8 models used example scenarios and 6 others provided some complexity and time analysis.  Of the 29 papers, we identified only
8 that provided in an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired were the model domain with 9 of the 29 models, enterprise workflows with 5 of 29, and mobile computing with 5 of 29.
\\
\\
\noindent
{\bf Conclusions}: The extensions to RBAC found by our systematic literature review fall into seven categoriies that upon closer inspection can be 
aligned under the single category of context. Two domains, healthcare and mobile computing, were idenitfied as predominant motivations behind the 
development of extensions to the RBAC model. Our literature review showed that the state of evaluation of RBAC model evaluation is limited with 1 
paper doing an evaluation comparing itself to the core RBAC model, 12 of the 29 models lacking any identifiable evaluation and 8 presenting example 
scenarios of their model in action, the predominant evaluation method.

### 2013-04-02 paper/sections/analysis.tex
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item What domains or scenarios serve as inspiration for these extensions?
\item Medical domain
\item Pervasive computing environments
\item Mobile devices
\item Large-scale organizations with many sub-departments
\item Enterprise, organization workflows
\end{itemize}

>>>>

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item What domains or scenarios serve as inspiration for these extensions?
\item Medical domain \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{motta03:contextual}, \cite{ni2010privacy}, \cite{damiani2007geo}, \cite{hansen2003spatial}, \cite{samuel07:spatio-temporal}, \cite{aich09:role}, \cite{zhou2007team}
\item Pervasive computing environments \cite{huang06:pervasive}, \cite{chen08:spatio-temporal}, \cite{ray07:spatio}
\item Web applications \cite{masoumzadeh2008purbac}
\item Mobile computing \cite{thein2011leveraging}, \cite{zou2009crbac}, \cite{chandran05:llt}, \cite{ray07:spatio}, \cite{aich09:role}
\item Large-scale organizations with many sub-departments \cite{yamazaki04:designing}, \cite{han08:extended}, \cite{yao2008task}
\item Enterprise, organization workflows \cite{cholewka00:acontext-sensitive}, \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{oh2003task}, \cite{joshi05:generalized}
\end{itemize}

### 2013-04-02 paper/sections/appendix.tex
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

\multicolumn{1}{ c }{\multirow{2}{*} {\rotatebox{90}{ Context } }} & 
\multicolumn{3}{ |l| }{A constraint based role based access control in the SECTET a model-driven approach, 2006 \cite{alam06:constraint}} \\ \cline{2-4}

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{A flexible content and context-based access control model for multimedia medical image database systems, 2001 \cite{tzelepi01:flexible}} \\ \cline{2-4}

@@

>>>>

\multicolumn{1}{ c }{\multirow{2}{*} {\rotatebox{90}{ Context } }} & 
\multicolumn{3}{ |l| }{A flexible content and context-based access control model for multimedia medical image database systems, 2001 \cite{tzelepi01:flexible}} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{{\multirow{1}{*}{\rotatebox{90}{ Const. }}}} &
\multicolumn{2}{ |l| }{A constraint based role based access control in the SECTET a model-driven approach, 2006 \cite{alam06:constraint}} \\[2.75ex] \cline{2-4}

### 2013-04-02 paper/sections/categorization.tex
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

\subsection{Research Question 1}

After deciding upon a set of primary sources, the authors noticed that each paper fell into one or more categorization's of extensions to RBAC.
The categorizations were extracted by looking at how they are defined, what are their characteristics and how do individual categories relate
to other categories, and encapsulated by the following research question:

What categorizations exist within RBAC extensions? \\

@@

The primary sources contained either within the title, or by multiple references within the body of the text a reference to a quantifier for the
type of extension their model was.  For example, the paper "Privacy-aware role-based access control" \cite{ni2010privacy} contained "privacy" in
the title leading to the papers categorization of Privacy.  By comparison, the paper "An extended RBAC model based on granular logic" \cite{jian2008extended}
does not contain a direct categorization in the title or model name, but in reading the body of the paper the authors assessed that this paper was 
context based.  The authors offer definitions for each categorization and their characteristics based on the primary sources and the English definitions
for each.

@@

  \item Constraint: Access control model in this category extends constraints, which are conditional restrictions on permissions of given roles. This constraint is either static and dynamic. For example, a doctor can modify any medical record for which the doctor is assigned as the designated primary care physician. This example describes doctor's permission with conditional restriction, "only when the doctor is assigned as the designated primary care physician".

@@

  \item Organizational: Organizational access control is concerned with access control associated with multiple organizations. Typically, users may have the same role name in different organizations, but may have different access privileges due to different local variations.

  \item Privacy: Access control model can be extended to describe privacy policies, which are legal statements or documents about disclose and management of personally identifiable information such as name, address, data of birth, etc.

  \item Resource: Access control model can be extended to handle any system resources (e.g., a file, printer, terminal, database record, etc) in a flexible way.

@@

  \item Temporal:  Temporal constraints are time-based constraints in specifying access
	control policies. For example, in organizations, 	periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should access required resources during the interval. 
	Temporal constraints can incorporate either on roles, user-role assignments, or role-permission assignments.

@@

\subsection{Relationships}

>>>>

\section{Categorization} \label{sec:categorization}

RQ1: What categories exist within RBAC extensions? \\

During the paper reading phase, the authors noted that a variety of themes were present amongst the extensions. 
For example, the paper ``Privacy-aware role-based access control'' \cite{ni2010privacy} brings in the notion of privacy directly in the title of the paper and the name of their model. 
Some papers presented this direct prouncement of their theme, while others were less obvious. 
Thus, the authors developed a set of guidelines to aide in determining a set of extension categories based upon observations during the paper reading phase. The guidelines were:

\begin{itemize}
\item Self Declaration - Does the title or name of the model classify itself?
\item Repetition of Phrase - Does the body of the paper present the same phrase repeatedly when discussing their model?
\item Self Assessment - Do the authors of the paper directly identify a quantifier for their paper within the body of the paper?
\item Simple - Can the category be represented by a single quantifier?
\end{itemize}

If we take the previous example above, the paper "Privacy-aware role-based access control" \cite{ni2010privacy} contained "privacy" in
the title and in the name of the model leading to the creation of the Privacy categorization and the subsequent placement of the paper under that category.
By comparison, the paper "An extended RBAC model based on granular logic" \cite{jian2008extended} does not contain a direct categorization in the title or model name, but in reading the body of the paper the authors assessed that this paper was context based.

The authors offer definitions for each category based on the data extracted from the primary sources and the English definitions for each word.

@@

@@

@@

  \item Constraint (Const): Access control model in this category extends constraints, which are conditional restrictions on permissions of given roles. This constraint is either static and dynamic. For example, a doctor can modify any medical record for which the doctor is assigned as the designated primary care physician. This example describes doctor's permission with conditional restriction, "only when the doctor is assigned as the designated primary care physician".

  \item Organizational (Org): Organizational access control is concerned with access control associated with multiple organizations. Typically, users may have the same role name in different organizations, but may have different access privileges due to different local variations.

  \item Privacy (Priv): Access control model can be extended to describe privacy policies, which are legal statements or documents about disclose and management of personally identifiable information such as name, address, data of birth, etc.

  \item Resource (Res): Access control model can be extended to handle any system resources (e.g., a file, printer, terminal, database record, etc) in a flexible way.

@@

  \item Temporal (Temp):  Temporal constraints are time-based constraints in specifying access
	control policies. For example, in organizations, 	periodic temporal durations are enforced while a
	specific role is permitted to conduct an action. Consider that part-time employee works only from 9:00 a.m. to 3:00 p.m.
	In such cases, the part-time employee role should access required resources during the interval. 
	Temporal constraints can incorporate either on roles, user-role assignments, or role-permission assignments.

@@

\subsection{Analysis and Discussion}

The 29 primary sources produced a set of 8 heirarchical categories. Table \ref{tab:categorization} summarizes each primary source under their assumed
category and further, displays the perceived heirarchy of the categories. 
The Spatial and Temporal categories were treated as subsets of the broader category of Spatio-Temporal since this category encompasses them individually and
the Spatio-Temporal category contained more primary sources that the Spatial or Temporal categories alone.

When looking across all categories, the authors noted that each category added some new features on top of the standard RBAC model that were domain specific.
These domain specific features were under the surface adding contextual relationships between the core user, permission and role entities.  Thus, the authors
concluded that all categories stemmed from the context category, of which some primary sources were already deemed direct members.

For example, in the case of Privacy the models added entities such as purpose binding to represent within the model data collected for one purpose should not used for another purpose without user consent {\cite{ni2010privacy}. 
While the new entity provided by the Privacy based models is inspired by domains such as healthcare where privacy is of legal concern, the underlying mechanism that drives purpose binding is providing context around making an access control decision. 
The system must take into account not just a static set of permissions a user has through their roles, but also the context of the data being accessed as that data relates to privacy policy. 
In the spatio-temporal models, a users location and the time of day are two factors that can be taken into account when activating a role or verifying a permission.  
The concepts of location and time are properties of the user and place specific contexts around the role and permission entities.

### 2013-04-02 paper/sections/intro.tex
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

Innovation and the use of RBAC into new domains has led to scenarios that, by some accounts, the standard RBAC model cannot handle.  For example, Ni, Q. and Bertino, E. et al \cite{ni2010privacy} state that RBAC is ``not designed to enforce privacy policies and barely meet privacy protection requirements`` with the introduction of privacy concerns in domains such as healthcare. Thus, a series of extensions to the standard RBAC model have appeared in the literature, each adding one or more features.  However, as researchers often develop their own specialized extended models of RBAC, their research cannot be generalized or compared with other research work appropriately.

Role based access control (RBAC) was first introduced in the 1990s when the National Institute for Standards and Technology (NIST) requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework proposed by Sandhu, et al \cite{sandhu1996role}.  The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2011 \footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}}. In an RBAC model, roles represent a group of users who are involved in a specific job function in an organization. RBAC assigns permissions of specific actions on resources to roles instead of individual users.  Therefore, in order to gain roles' permission on specific resources, users acquire appropriate roles first.  RBAC is a generalized access control approach used for various applications including web services, database applications, and healthcare applications.  RBAC has advantages in maintaining and managing organization's security policies.  For example, if a user is to access manager role's resources within a given organization, security policy administrators simply add the user to be associated with the manager role.

@@

The rest of the paper is organized as follows.  Section 2 covers background and the core RBAC model. Section 3 lays out the process used in conducting the review. Section 4 discusses the categorization break down of the papers. Section 5 analyzes the rest of the research questions.  Section 6 contains the final conclusions.

>>>>

Innovation and the use of role based access control (RBAC) into new domains has led to scenarios that, by some accounts, the standard RBAC model cannot handle \cite{kuhn2010adding}.  
For example, Ni, Q. et al \cite{ni2010privacy} state that RBAC is ``not designed to enforce privacy policies and barely 
meet privacy protection requirements`` with the introduction of privacy concerns in domains such as healthcare. 
Since the introduciton of the NIST RBAC standard in 2000, extension models to the RBAC standard have appeared in the literature, each adding one or more features onto the core.  
A cursory examination of the extension model papers does not provide good comparisons. Further, no generalizations among models have been made which inherently defies the reasoning behind
creating a standard.

The RBAC standard was first introduced in the 1990s when the National Institute for Standards and Technology (NIST) 
requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework 
proposed by Sandhu, et al \cite{sandhu1996role}.  In 2004, this standard was adopted as ANSI/INCITS 359-2004.  The development of a standard was inspired by an economic impact study done 
during the 1990s and later confirmed in 2011 \footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}}
that showed the cost savings of RBAC in implementation and maintenance.
RBAC is a generalized access control approach used for various applications including web services, 
database applications, and healthcare applications.  RBAC has advantages in maintaining and managing organization's security policies.  
For example, if a user is to access manager role's resources within a given organization, security policy administrators simply add the user to be associated with the manager role.
In an RBAC model, roles represent a group of users who are involved in a specific job function in an organization. RBAC assigns permissions 
of specific actions on resources to roles instead of individual users. Therefore, in order to gain a roles' permission on specific resources, 
users acquire appropriate roles first.

@@

The rest of the paper is organized as follows.  Section 2 covers background and the core RBAC model. Section 3 lays out the process used in conducting the review. Sections 4-9 present the analysis and discussion of the research questions. Section 10 contains the final conclusions.

### 2013-04-02 paper/sections/process.tex
Breaks each research quesiton into it's own major section.
Adding adjustments to abstract and introduction with focus on standards. Re-ordeirng the categories and adding short names for categories as seen in the table. Adding analysis and discussion section for categories.

\begin{itemize}
\item Step 1: Development of a search strategy
\item Step 2: Elimination of papers based on title criteria
\item Step 3: Elimination of papers based on abstract criteria
\item Step 4: Elimination of papers based on content and matching to elimination criteria
\end{itemize}

@@

\begin{enumerate}
\item Call to search engine with current search position and current search term.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria:
	\begin{itemize}
	\item If the size of the result set is greater than or equal to 1000 then stop.
	\item If the last ten results did not contain the search term phrase within the tite then stop.
	\end{itemize}
\item If stopping criteria not met, increment search position and go back to step one.
\end{enumerate}

@@

TODO: Add elimination stats.

@@

\begin{itemize}
\item The two first authors independently classified papers as relevant, irrelevant or uncertain based off elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant, or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.
\end{itemize}

@@

\begin{itemize}
\item Did the title contain a reference to 'role based access control' or 'RBAC'? 
\item Did the title contain a reference to 'model'?
\end{itemize}

@@

\begin{itemize}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention either an implementation, evaluation or domain for their model?
\end{itemize}

@@

\begin{enumerate}
\item Does this model extend the core model? (Exclusion)
\item Do the researchers give evidence that RBAC needs extension? (Inclusion)
\item Was the paper and subsequent model inspired by a real world example?  (Conditional Inclusion)
\item Did the researchers offer any evaluation of the proposed model? If yes, how did they do one? If no, why? (Conditional Inclusion)
\item Did the authors implement their model? (Inclusion)
\end{enumerate}

@@

 & \multicolumn{1}{ c| }{} & \textbf{Title} & \textbf{Abstract} & \textbf{Paper Body} \\ \hline

@@

>>>>

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Step 1: Development of a search strategy
\item Step 2: Elimination of papers based on title criteria
\item Step 3: Elimination of papers based on abstract criteria
\item Step 4: Elimination of papers based on content and matching to elimination criteria
\end{itemize}

@@

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Call to search engine with current search position and current search term.
\item Parse results and extract paper title, authors and year of publication.
\item Compare results against stopping criteria:
	\begin{itemize}
	\item If the size of the result set is greater than or equal to 1000 then stop.
	\item If the last ten results did not contain the search term phrase within the tite then stop.
	\end{itemize}
\item If stopping criteria not met, increment search position and go back to step one.
\end{enumerate}

@@

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The two first authors independently classified papers as relevant, irrelevant or uncertain based off elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant, or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.
\end{itemize}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Did the title contain a reference to 'role based access control' or 'RBAC'? 
\item Did the title contain a reference to 'model'?
\end{itemize}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention either an implementation, evaluation or domain for their model?
\end{itemize}

@@

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Does this model extend the core model? (Exclusion)
\item Do the researchers give evidence that RBAC needs extension? (Inclusion)
\item Was the paper and subsequent model inspired by a real world example?  (Conditional Inclusion)
\item Did the researchers offer any evaluation of the proposed model? If yes, how did they do one? If no, why? (Conditional Inclusion)
\item Did the authors implement their model? (Inclusion)
\end{enumerate}

@@

 & \multicolumn{1}{ c| }{} & \textbf{Title} & \textbf{Abstract} & \textbf{Content} \\ \hline

@@

\subsection{Extraction}

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round, and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for the RBAC model extensions and what papers feel into what categories.
